Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-03-25 SIR_manuscript.tex
replace eta by 1-theta in Appendix (still can be simplified more)

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $X_n$, $X_p$ and $X_c$ for $X \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta\_w=0$ and $\eta\_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

@@

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta\_c \ll \eta\_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $w_I \gg w_S$).

@@

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are \ali{all $S\_u \& S\_n$ should be $S\_u^* \& S\_n^*$}
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u&\eta\_w\,S\_u&\eta\_w\,S\_u&\eta\_c\,S\_u\\
S\_n&\eta\_w\,S\_n&\eta\_w\,S\_n&\eta\_c\,S\_n\\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right], \\
  V =&
 \left[ \begin {array}{cccc}  
F_I+\gamma&-\omega&0&0\\
-(1-p_I)F_I&\omega+\gamma&0&0\\
-p_I F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right], \text{thus}\\
V^{-1} =&
\left[ \begin {array}{cccc}
\frac {\omega+\gamma}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&\frac {\omega}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&0&0\\
\noalign{\medskip}
\frac{(1-p_I) F_I}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&
\frac{F_I+\gamma}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&0&0\\
\noalign{\medskip}
\frac{p_I F_I}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&
\frac{\omega p_I F_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I)(\omega+\gamma)}& \frac{1}{\omega+\gamma}&0 \\
\noalign{\medskip}
\frac{\omega\,F_I\,p_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I) \gamma}& 
\frac{\omega^2\,F_I\,p_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I)(\omega+\gamma) \gamma}&
\frac{\omega}{(\omega+\gamma) \gamma}&
\frac{1}{\gamma}
\end {array} \right].
\end{align}

@@

% %%%%%%%
\subsection{Taylor Approximation of $\Rnum$ at $\rho=0$ }
The basic reproduction number, $\Rnum$, close to $\rho=0$ can be approximated linearly in $\rho$ by using Taylor approximation. it follows
\begin{equation}
\label{eq:R0appr}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(\eta\_w-1)(\gamma w_S+\omega w_I) + (\eta\_c -1)p_I w_I \omega^2 \Big) + \mathcal{O}(\rho^2).
\end{equation}

Also, to analyze the influence of $\omega$ on $\Rnum$, the approximation \eqref{eq:R0appr} was used. Then it is straight forward to have
\begin{equation}
\label{Rom}
\partial{\Rnum}/\partial{\omega}=  \frac{-\beta \rho}{\gamma w_S\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c),
\end{equation}
where $a=(\eta\_w-1)w_I-(\eta\_c-1)p_I w_I = ((s-p_I)\eta\_c + (p_I-1)) w_I$, $b=2(\eta\_w-1)\gamma w_S$ and $c=(\eta\_w-1)\gamma^2 w_S$.
Given that $0 \leq \eta\_c\leq \eta\_w \leq 1 $, one can easily derive $b\leq 0$ and $c \leq 0$.

Note that in general, the necessary and sufficient condition for $a \geq 0$ is $(s-p_I) \eta\_c \geq (1-p_I)$, where $s=\frac{\eta\_w}{\eta\_c} \geq 1$.

In the case of ``perfect isolation'', i.e., when $\eta\_w=0$ and consequently $\eta\_c=0$, it is straightforward to see that $a \leq 0$, $b<0$ and $c<0$. Thus, $\partial{\Rnum}/\partial{\omega} \geq 0$. 
As an example, in the case of a very accurate testing regime,  i.e., $P_I=1$, $a \geq 0$ is achieved. If $a\geq 0$, the quadratic expression in \eqref{Rom}, has Real roots. Assuming that $\omega_1<0$ and $\omega_2>0$ be the roots of the quadratic expression in $\partial{\Rnum}/\partial{\omega}$. Thus, $\partial{\Rnum}/\partial{\omega}>0$ for $0<\omega<\omega_2$ and  $\partial{\Rnum}/\partial{\omega}<0$ for $\omega>\omega_2$.

% %%%%%%%
\subsection{rate of returning tests} 
\ali{needs editig}
The linearization of $\Rnum$ around $\rho=0$ is
\begin{equation}\label{linearization}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(\eta\_w-1)(\gamma w_S+\omega w_I) + (\eta\_c -1)p_I w_I \omega^2 \Big). 
\end{equation}
So when $\rho \approx 0$ we have $$\partial{\Rnum}/\partial{\omega} \approx  \frac{-\beta \rho}{\gamma w_S\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c),$$
where $a=(\eta\_w-1)w_I-(\eta\_c-1)p_I w_I$, $b=2(\eta\_w-1)\gamma w_S$ and $c=(\eta\_w-1)\gamma^2 w_S$.

@@

Note that the zeroes are real if and only if $Q < 1$. Note that have $\eta\_c < \eta\_w$, so if $P_i \approx 1$, we will have $Q < 0 < 1$. Thus, if we assume near-perfect test sensitivity, $\omega_1$ and $\omega_2$ will be real.

@@

Consider a test that allows us to test at rate $\rho_1$ and has sensitivity $P_{i,1}$, and another test that allows us to test at $\rho_2$ and has sensitivity $P_{i,2}$. Suppose that $\rho_1 > \rho_2$. Recall that the linearization of $\Rnum$ around $\rho \approx 0$ is given by $$\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(\eta\_w-1)(\gamma w_S+\omega w_I) + (\eta\_c -1)p_I w_I \omega^2 \Big).$$

@@

\begin{align}\label{eq:rho1vsrho2}
    &\rho_1\left(\gamma(\eta\_w-1)(\gamma w_S + \omega w_I) + (\eta\_c-1)p_{I,1}w_I\omega^2\right) - \rho_2\left(\gamma(\eta\_w-1)(\gamma w_S + \omega w_I) + (\eta\_c-1)p_{I, 2}w_I\omega^2\right) > 0 \nonumber \\
    &\vdots \nonumber \\
    &\frac{\rho_2P_{i, 2}-\rho_1P_{i, 1} }{\rho_1-\rho_2} > \frac{1-\eta\_w}{1-\eta\_c}\cdot \frac{\gamma(\gamma w_S + \omega w_I)}{\omega^2 w_I}
\end{align}

>>>>

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $X_n$, $X_p$ and $X_c$ for $X \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\theta\_w=1$ and $\theta\_c=1$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

@@

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\theta\_c \gg \theta\_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $w_I \gg w_S$).

@@

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are \ali{all $S\_u \& S\_n$ should be $S\_u^* \& S\_n^*$}
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u&(1-\theta\_w)\,S\_u&(1-\theta\_w)\,S\_u&(1-\theta\_c)\,S\_u\\
S\_n&(1-\theta\_w)\,S\_n&(1-\theta\_w)\,S\_n&(1-\theta\_c)\,S\_n\\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right], \\
  V =&
 \left[ \begin {array}{cccc}  
F_I+\gamma&-\omega&0&0\\
-(1-p_I)F_I&\omega+\gamma&0&0\\
-p_I F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right], \text{thus}\\
V^{-1} =&
\left[ \begin {array}{cccc}
\frac {\omega+\gamma}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&\frac {\omega}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&0&0\\
\noalign{\medskip}
\frac{(1-p_I) F_I}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&
\frac{F_I+\gamma}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&0&0\\
\noalign{\medskip}
\frac{p_I F_I}{\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I}&
\frac{\omega p_I F_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I)(\omega+\gamma)}& \frac{1}{\omega+\gamma}&0 \\
\noalign{\medskip}
\frac{\omega\,F_I\,p_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I) \gamma}& 
\frac{\omega^2\,F_I\,p_I}{(\omega\,\gamma+\gamma F_I+\gamma^2+\omega\, F_I p_I)(\omega+\gamma) \gamma}&
\frac{\omega}{(\omega+\gamma) \gamma}&
\frac{1}{\gamma}
\end {array} \right].
\end{align}

@@

% %%%%%%%
\subsection{Taylor Approximation of $\Rnum$ at $\rho=0$ }
The basic reproduction number, $\Rnum$, close to $\rho=0$ can be approximated linearly in $\rho$ by using Taylor approximation. it follows
\begin{equation}
\label{eq:R0appr}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big) + \mathcal{O}(\rho^2).
\end{equation}

Also, to analyze the influence of $\omega$ on $\Rnum$, the approximation \eqref{eq:R0appr} was used. Then it is straight forward to have
\begin{equation}
\label{Rom}
\partial{\Rnum}/\partial{\omega}=  \frac{-\beta \rho}{\gamma w_S\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c),
\end{equation}
where $a=(-\theta\_w)w_I-(-\theta\_c)p_I w_I = ((s-p_I)(1-\theta\_c) + (p_I-1)) w_I$, $b=2(-\theta\_w)\gamma w_S$ and $c=(-\theta\_w)\gamma^2 w_S$.
Given that $0 \leq \theta\_w \leq \theta\_c \leq 1 $, one can easily derive $b\leq 0$ and $c \leq 0$.

Note that in general, the necessary and sufficient condition for $a \geq 0$ is $(s-p_I) (1-\theta\_c) \geq (1-p_I)$, where $s=\frac{(1-\theta\_w)}{(1-\theta\_c)} \geq 1$.

In the case of ``perfect isolation'', i.e., when $\theta\_w=1$ and consequently $\theta\_c=1$, it is straightforward to see that $a \leq 0$, $b<0$ and $c<0$. Thus, $\partial{\Rnum}/\partial{\omega} \geq 0$. 
As an example, in the case of a very accurate testing regime,  i.e., $P_I=1$, $a \geq 0$ is achieved. If $a\geq 0$, the quadratic expression in \eqref{Rom}, has Real roots. Assuming that $\omega_1<0$ and $\omega_2>0$ be the roots of the quadratic expression in $\partial{\Rnum}/\partial{\omega}$. Thus, $\partial{\Rnum}/\partial{\omega}>0$ for $0<\omega<\omega_2$ and  $\partial{\Rnum}/\partial{\omega}<0$ for $\omega>\omega_2$.

% %%%%%%%
\subsection{rate of returning tests} 
\ali{needs editig}
The linearization of $\Rnum$ around $\rho=0$ is
\begin{equation}\label{linearization}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big). 
\end{equation}
So when $\rho \approx 0$ we have $$\partial{\Rnum}/\partial{\omega} \approx  \frac{-\beta \rho}{\gamma w_S\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c),$$
where $a=(-\theta\_w)w_I-(-\theta\_c)p_I w_I$, $b=2(-\theta\_w)\gamma w_S$ and $c=(-\theta\_w)\gamma^2 w_S$.

@@

Note that the zeroes are real if and only if $Q < 1$. Note that have $\theta\_c > \theta\_w$, so if $p_I \approx 1$, we will have $Q < 0 < 1$. Thus, if we assume near-perfect test sensitivity, $\omega_1$ and $\omega_2$ will be real.

@@

Consider a test that allows us to test at rate $\rho_1$ and has sensitivity $P_{i,1}$, and another test that allows us to test at $\rho_2$ and has sensitivity $P_{i,2}$. Suppose that $\rho_1 > \rho_2$. Recall that the linearization of $\Rnum$ around $\rho \approx 0$ is given by $$\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big).$$

@@

\begin{align}\label{eq:rho1vsrho2}
    &\rho_1\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I,1}w_I\omega^2\right) - \rho_2\left(\gamma(-\theta\_w)(\gamma w_S + \omega w_I) + (-\theta\_c)p_{I, 2}w_I\omega^2\right) > 0 \nonumber \\
    &\vdots \nonumber \\
    &\frac{\rho_2P_{i, 2}-\rho_1P_{i, 1} }{\rho_1-\rho_2} > \frac{\theta\_w}{\theta\_c}\cdot \frac{\gamma(\gamma w_S + \omega w_I)}{\omega^2 w_I}
\end{align}
